# 12. IP weighting and marginal structural models {-} 

<!--## Concepts and points{-}

- Confounder
-->

## Estimating IP weights via modeling{-} 

### Definition{-}

An individual's IP weight depends on her values of treatment $A$ and covariate $L$. Since the denominator of the weight for each individual is the conditional density evaluated at the individual's own values of $A$ and $L$, it can be expressed as the conditional density evaluated at the random arguments $A$ and $L$, that is, as $f(A|L)$. Therefore, we write the IP weights as $1/f(A|L)$. 

- Potential outcome mean for treatment level $A=a$: $E(Y^{a})$
- Standardized mean for treatment level $A=a$: $\sum_{l}E(Y|A=a, L=l)P(L=l)$
- IP weighted mean of $Y$ for treatment level $A=a$: $E\left(\frac{I(A=a)Y}{f(A|L)}\right)$

### Equivalence of IP weighting and standardization{-}
Under the positivity assumption, we have 
\begin{equation*}
\begin{split}
E\left(\frac{I(A=a)Y}{f(A|L)}\right)=E_{A, L}\left( E_{Y}(\frac{I(A=a)}{f(A|L)}Y|A, L)\right)=E_{A, L}\left( E_{Y}(Y|A, L)\frac{I(A=a)}{f(A|L)}\right) \\
= E_{L}\left(E_{A}\left( E_{Y}(Y|A, L)\frac{I(A=a)}{f(A|L)}|L\right)\right)=E_{L}\left( E_{Y}(Y|A=a, L)|L\right) \\
=\sum_{l}E(Y|A=a, L=l)P(L=l)
\end{split}
\end{equation*}

### Equivalence of potential outcome mean, standardized mean and IP weighted mean{-} 
- First, we show $E(Y^a)=\sum_{l}E(Y|A=a, L=l)P(L=l)$ under conditional exchangeability, positivity, and consistency. 

\begin{equation*}
\begin{split}
E(Y^{a})=\sum_{l}E(Y^a|L=l)P(L=l) \\
=\sum_{l}E(Y^a|A=a, L=l)P(L=l)=\sum_{l}E(Y|A=a, L=l)P(L=l), 
\end{split}
\end{equation*}
where the second equality is by conditional exchangeability and positivity, and the third by consistency. 

- Second, we show $E(Y^a)=E\left( \frac{I(A=a)}{f(A|L)}Y \right)$ under positivity, conditional exchangeability, and consistency. 

\begin{equation*}
\begin{split}
E\left(\frac{I(A=a)}{f(A|L)}Y\right)= E\left( \frac{I(A=a)}{f(A|L)}Y^a \right)=E\left(  E\left(\frac{I(A=a)}{f(A|L)}Y^{a}|L\right) \right) \\
= E\left(E\left(\frac{I(A=a)}{f(A|L)}|L\right)E\left(Y^a|L\right) \right)=E\left(1\times E\left(Y^{a}|L\right)  \right)=E(Y^a),
\end{split}
\end{equation*}
where the first equality is by consistency, the third by conditional exchangeability. Furthermore, we need positivity to well define the IP weights. 

### What does IP weighting mean? {-}

IP weighting creates a pseudo-population in which the arrow from the covariates $L$ to the treatment $A$ is removed (ther is no association between covariates $L$ and the treatment $A$). The pseudo-population has the following two properties:
- $A$ and $L$ are statistically independent
- the mean $E_{ps}(Y|A=a)$ in the pseudo-population equals the standardized mean $\sum_{l}E(Y|A=a, L=l)P(L=l)$ in the actual population under positivity

### Horvitz-Thompson estimator and Hajek estimator{-}

- The IP weighted mean can be consistently estimated by Horvitz-Thompson estimator $\widehat{E}(\frac{I(A=a)}{f(A|L)}Y)$

or 

- the Hajek estimator $\frac{\widehat{E}(\frac{I(A=a)}{f(A|L)}Y)}{\widehat{E}(\frac{I(A=a)}{f(A|L)})}$, which is an unbiased estimator of $\frac{E(\frac{I(A=a)}{f(A|L)}Y)}{E(\frac{I(A=a)}{f(A|L)})}$. Note under positivity, $\frac{E(\frac{I(A=a)}{f(A|L)}Y)}{E(\frac{I(A=a)}{f(A|L)})}=E(\frac{I(A=a)}{f(A|L)}Y)$. In practice, the Hajek estimator is preferred because it is guaranteed to lie between 0 and 1 for dichotomous $Y$. 


## Stablized IP weights {-}

- Several facts from mathematics: under positivity, conditional exchangeability, and consistency, 
\begin{equation*}
\frac{E(\frac{pI(A=a)}{f(A|L)}Y)}{E(\frac{pI(A=a)}{f(A|L)})}=E(Y^a),
\end{equation*}
and 
\begin{equation*}
\frac{E(\frac{g(A)I(A=a)}{f(A|L)}Y)}{E(\frac{g(A)I(A=a)}{f(A|L)})}=E(Y^a),
\end{equation*}
where g(A) is any function of $A$ that is not a function of $L$. Usually we take $g(A)=f(A)$. 

- The IP weights $W^{A}=1/f(A|L)$ are referred to as nonstabilized weights, and the IP weights $SW^{A}=f(A)/f(A|L)$ are referred to as stabilized weights. The stablizing factor $f(A)$ in the numerator is responsible for the narrower range of the $f(A)/f(A|L)$ weights.   

- More on IPW: http://www.rebeccabarter.com/blog/2017-07-05-ip-weighting/

## Marginal structural models {-}

- Examples of marginal structural mean models: $E(Y^a)=\beta_0+\beta_1a$ or $E(Y^a)=\beta_0+\beta_1a+\beta_2a^2$













## Real data analysis{-}

### Background{-}
- NHEFS data (National Health and Nutrition Examination Survey Data I Epidemiologic Follow-up Study)
- Goal: estimate the effect of smoking cessation on weight gain 
- $A=1$: if cigarette smokers reported having quit smoking before the follow-up visit
- $Y$: the body weight at the follow-up visit minus the body weight at the baseline visit
- $E(Y^{a=1})$: mean weight gain that would have been observed if all individuals in the population had quit smoking before the follow-up visit
- $E(Y^{a=0})$: mean weight gain that would have been observed if all individuals in the population had not quit smoking
- Average causal effect: $E(Y^{a=1})-E(Y^{a=0})$


```{r}
library(here) # for path 
library(readxl) # for reafing excel files 
library(tidyverse)
library(hrbrthemes)
library(viridis)
library(geepack) # for GEE
# copy from https://www.r-graph-gallery.com/89-box-and-scatter-plot-with-ggplot2.html
```

### Input dataset{-}

```{r}
nhefs=read_excel(here("data", "NHEFS.xls"))
head(nhefs)
```

### Ignore subjects with missing values for weight in 1982{-} 
- ```wt82```: weight in 1982
- ```qsmk```: quit smoking between 1st questionnaire and 1982 (Yes:1; No:0)

```{r}
nhefs.nmv=nhefs[which(!is.na(nhefs$wt82)), ]
nhefs.nmv$qsmk=as.factor(nhefs.nmv$qsmk)
```


### Compare the treatment group and the control group{-} 
- ```age```
```{r}
summary(nhefs.nmv[which(nhefs.nmv$qsmk==0), ]$age)
summary(nhefs.nmv[which(nhefs.nmv$qsmk==1), ]$age)

nhefs.nmv %>% ggplot( aes(x=qsmk, y=age, fill=qsmk)) +
      geom_boxplot()+ 
      scale_fill_viridis(discrete = TRUE, alpha=0.6) +
      geom_jitter(color="black", size=0.4, alpha=0.9) +
      theme_ipsum() +
      theme(legend.position="none", plot.title = element_text(size=11)) +
      ggtitle("Boxplot") +
      xlab("qsmk")
```

- ```wt71```

```{r}
summary(nhefs.nmv[which(nhefs.nmv$qsmk==0), ]$wt71)
summary(nhefs.nmv[which(nhefs.nmv$qsmk==1), ]$wt71)


nhefs.nmv %>% ggplot( aes(x=qsmk, y=wt71, fill=qsmk)) +
      geom_boxplot()+ 
      scale_fill_viridis(discrete = TRUE, alpha=0.6) +
      geom_jitter(color="black", size=0.4, alpha=0.9) +
      theme_ipsum() +
      theme(legend.position="none", plot.title = element_text(size=11)) +
      ggtitle("Boxplot") +
      xlab("qsmk")





```

- ```smokeintensity```: number of cigarettes smoked per day in 1971
    
```{r}

summary(nhefs.nmv[which(nhefs.nmv$qsmk==0), ]$smokeintensity)
summary(nhefs.nmv[which(nhefs.nmv$qsmk==1), ]$smokeintensity)


nhefs.nmv %>% ggplot( aes(x=qsmk, y=smokeintensity, fill=qsmk)) +
      geom_boxplot()+ 
      scale_fill_viridis(discrete = TRUE, alpha=0.6) +
      geom_jitter(color="black", size=0.4, alpha=0.9) +
      theme_ipsum() +
      theme(legend.position="none", plot.title = element_text(size=11)) +
      ggtitle("Boxplot") +
      xlab("qsmk")







```  

- ```smokeyrs```: years of smoking 
```{r}
summary(nhefs.nmv[which(nhefs.nmv$qsmk==0), ]$smokeyrs)
summary(nhefs.nmv[which(nhefs.nmv$qsmk==1), ]$smokeyrs)


nhefs.nmv %>% ggplot( aes(x=qsmk, y=smokeyrs, fill=qsmk)) +
      geom_boxplot()+ 
      scale_fill_viridis(discrete = TRUE, alpha=0.6) +
      geom_jitter(color="black", size=0.4, alpha=0.9) +
      theme_ipsum() +
      theme(legend.position="none", plot.title = element_text(size=11)) +
      ggtitle("Boxplot") +
      xlab("qsmk")

```   
   
- sex

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$sex)

prop.table(table(nhefs.nmv$qsmk, nhefs.nmv$sex), margin=1) # row-wise

nhefs.nmv$sex=as.factor(nhefs.nmv$sex)
nhefs.nmv%>%ggplot(aes(x = qsmk, fill = sex)) + 
  geom_bar(position = "dodge")



```


- race: 0 if white, 1 if black or other in 1971

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$race)

prop.table(table(nhefs.nmv$qsmk, nhefs.nmv$race), margin=1) # row-wise

nhefs.nmv$race=as.factor(nhefs.nmv$race)
nhefs.nmv%>%ggplot(aes(x = qsmk, fill = race)) + 
  geom_bar(position = "dodge")



```

- education (AMOUNT OF EDUCATION BY 1971): 1 if 8TH GRADE OR LESS, 2 if HS DROPOUT, 3 if HS, 4 if COLLEGE DROPOUT, 5 if COLLEGE OR MORE

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$education)

prop.table(table(nhefs.nmv$qsmk, nhefs.nmv$education), margin=1) # row-wise

nhefs.nmv$education=as.factor(nhefs.nmv$education)
nhefs.nmv%>%ggplot(aes(x = qsmk, fill = education)) + 
  geom_bar(position = "dodge")



```


- exercise: 0 if much exercise,1 if moderate exercise,2 if little or no exercise

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$exercise)

prop.table(table(nhefs.nmv$qsmk, nhefs.nmv$exercise), margin=1) # row-wise

nhefs.nmv$exercise=as.factor(nhefs.nmv$exercise)
nhefs.nmv%>%ggplot(aes(x = qsmk, fill = exercise)) + 
  geom_bar(position = "dodge")



```

- active: 0 if very active, 1 if moderately active, 2 if inactive

```{r}
table(nhefs.nmv$qsmk, nhefs.nmv$active)

prop.table(table(nhefs.nmv$qsmk, nhefs.nmv$active), margin=1) # row-wise

nhefs.nmv$active=as.factor(nhefs.nmv$active)
nhefs.nmv%>%ggplot(aes(x = qsmk, fill = active)) + 
  geom_bar(position = "dodge")



```

### Estimating IP weights via modeling {-}

- Step 1: Estimate $f(A|L)$ by logistic regression and let $\widehat{W}_i=\frac{1}{\widehat{P}(A|L)}$
- Step 2: Estimate the IP weighted mean $E\left(\frac{I(A=a)}{f(A|L)}Y\right)$ by Hajek estimator $\frac{\sum_{i}\widehat{W}_iY_i}{\sum_{i}\widehat{W}_i}$, where the sum is over all subjects with $A_i=a$

```{r}
## run logistic regression
logfit<-glm(
  qsmk~sex+race+age+I(age^2)+education+smokeintensity+I(smokeintensity^2)+
  smokeyrs+I(smokeyrs^2)+exercise+active+wt71+I(wt71^2), 
  family=binomial(),
  data=nhefs.nmv
)

summary(logfit)



```


```{r}
## calculate estimated probability
p.qsmk.obs=ifelse(nhefs.nmv$qsmk==0, 
                  1-predict(logfit, type="response"), 
                  predict(logfit, type="response")   ## this predicts the conditional probability of treatment 1
                  )
nhefs.nmv$ipw=1/p.qsmk.obs
summary(nhefs.nmv$ipw)

```

```{r}
## run weighted least squares 

wls=geeglm(wt82_71~qsmk, 
           data=nhefs.nmv,
           weights=ipw,
           id=seqn,
           corstr = "independence"
           )
summary(wls)

## build 95% confidence intervals for coefficients
## use the robust variance estimator, see more details in page 152 of the book
theta=coef(wls)
se=coef(summary(wls))[,2]
ci.lower=theta-qnorm(0.975)*se
ci.upper=theta+qnorm(0.975)*se
cbind(theta, ci.lower, ci.upper)

```


```{r}
## estimate average causal effect

## A=1
y1=nhefs.nmv[nhefs.nmv$qsmk==1, ]$wt82_71
w1=nhefs.nmv[nhefs.nmv$qsmk==1, ]$ipw

## A=0
y0=nhefs.nmv[nhefs.nmv$qsmk==0, ]$wt82_71
w0=nhefs.nmv[nhefs.nmv$qsmk==0, ]$ipw

ate=sum(y1*w1)/sum(w1)-sum(y0*w0)/sum(w0)

ate

```


```{r}
## association of sex and qsmk in the original population
xtabs(~nhefs.nmv$sex+nhefs.nmv$qsmk)


## no association between sex and qsmk in the pseudo-population

xtabs(nhefs.nmv$ipw~nhefs.nmv$sex+nhefs.nmv$qsmk)

```


```{r}
## check for positivity (white women)

xtabs(~nhefs.nmv$age[nhefs.nmv$race==0&nhefs.nmv$sex==1]+
       nhefs.nmv$qsmk[nhefs.nmv$race==0&nhefs.nmv$sex==1])

```


### Stablized IP weights {-}

```{r}
## estimate the denominator and numerator of the stablized IP weights 
denom.logfit<-glm(
  qsmk~sex+race+age+I(age^2)+education+smokeintensity+I(smokeintensity^2)+
  smokeyrs+I(smokeyrs^2)+exercise+active+wt71+I(wt71^2), 
  family=binomial(),
  data=nhefs.nmv
)

summary(denom.logfit)

numer.logfit<-glm(qsmk~1, 
                  family=binomial(),
                  data=nhefs.nmv)
summary(numer.logfit)

pd.qsmk=predict(denom.logfit, type="response")
pn.qsmk=predict(numer.logfit, type="response")

nhefs.nmv$sw=ifelse(nhefs.nmv$qsmk==0, ((1-pn.qsmk)/(1-pd.qsmk)), (pn.qsmk/pd.qsmk))





```


```{r}
## weighted least squares with the stabilized IP weights 

wls.sw=geeglm(wt82_71~qsmk,
              data=nhefs.nmv,
              weights=sw,
              id=seqn,
              corstr="independence"
              )
summary(wls.sw)

theta=coef(wls.sw)
se=coef(summary(wls.sw))[,2]
ci.lower=theta-qnorm(0.975)*se
ci.upper=theta+qnorm(0.975)*se
cbind(theta, ci.lower, ci.upper)






```


```{r}
## no association between sex and qsmk in the pseudo population

xtabs(nhefs.nmv$sw~nhefs.nmv$sex+nhefs.nmv$qsmk)



```


```{r}
## estimate average causal effect

## A=1
y1=nhefs.nmv[nhefs.nmv$qsmk==1, ]$wt82_71
w1=nhefs.nmv[nhefs.nmv$qsmk==1, ]$sw

## A=0
y0=nhefs.nmv[nhefs.nmv$qsmk==0, ]$wt82_71
w0=nhefs.nmv[nhefs.nmv$qsmk==0, ]$sw

ate=sum(y1*w1)/sum(w1)-sum(y0*w0)/sum(w0)

ate

```


### Marginal structural models {-}

- Estimating stabilized weights 
  - Approaches in the book: we need to estimate the stabilized weights $SW^A=f(A)/f(A|L)$. For a continuous treatment $A$, $f(A|L)$ is a probability density function. We assume that the density $f(A|L)$ is normal with mean $\mu_L=E(A|L)$ and constant variance $\sigma^2$. Then we use a linear regression model to estimate the mean $E(A|L)$ and variance of residuals $\sigma^2$ for all combinations of values of $L$. We also assume the density $f(A)$ is normal. 
  
  
```{r}
## analysis restricted to subjects reporting <=25 cig/day at baseline
nhefs.nmv.s=subset(nhefs.nmv, smokeintensity<=25)

## estimating denominator in ip weights 
den.fit.obj=lm(
  smkintensity82_71~sex+race+age+I(age^2)+
    education+smokeintensity+I(smokeintensity^2)+
    smokeyrs+I(smokeyrs^2)+exercise+active+wt71+I(wt71^2),
  data=nhefs.nmv.s
)

p.den=predict(den.fit.obj, type="response")
dens.den=dnorm(nhefs.nmv.s$smkintensity82_71, 
                mean=p.den,
                sd=summary(den.fit.obj)$sigma)


## estimating numerator in ip weights 
num.fit.obj=lm(smkintensity82_71~1, data=nhefs.nmv.s)
p.num=predict(num.fit.obj, type="response")
dens.num=dnorm(nhefs.nmv.s$smkintensity82_71, 
               p.num, 
               summary(num.fit.obj)$sigma)

nhefs.nmv.s$sw.a=dens.num/dens.den
summary(nhefs.nmv.s$sw.a)

```

  
```{r}
msm.sw.cont=geeglm(wt82_71~smkintensity82_71+I(smkintensity82_71*smkintensity82_71),
data=nhefs.nmv.s,
weights=sw.a,
id=seqn,
corstr="independence")

summary(msm.sw.cont)

beta=coef(msm.sw.cont)
SE=coef(summary(msm.sw.cont))[,2]
lcl=beta-qnorm(0.975)*SE
ucl=beta+qnorm(0.975)*SE
cbind(beta, lcl, ucl)
```





